% main.tex — P01: Persistent Homology of UK Socioeconomic Life-Course Trajectories
% arXiv preprint / Sociological Methodology submission
%
% Build: latexmk -pdf main.tex  (or see Makefile)

\documentclass[12pt, a4paper]{article}

% --- Encoding & fonts ---
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{mathpazo}           % Palatino for body text
\usepackage{inconsolata}        % monospace

% --- Page geometry ---
\usepackage[margin=1in]{geometry}
\usepackage{setspace}
\doublespacing                  % journal requirement

% --- Mathematics ---
\usepackage{amsmath, amssymb, amsthm}

% --- Tables ---
\usepackage{booktabs}
\usepackage{array}
\usepackage{multirow}
\usepackage{longtable}

% --- Graphics ---
\usepackage{graphicx}
\graphicspath{{../}}
\usepackage[font=small, labelfont=bf]{caption}
\usepackage{subcaption}

% --- Hyperlinks (load last among basics) ---
\usepackage[colorlinks=true, linkcolor=blue!60!black, citecolor=blue!60!black, urlcolor=blue!60!black]{hyperref}

% --- Bibliography ---
\usepackage[round, authoryear]{natbib}
\bibliographystyle{apalike}

% --- Misc ---
\usepackage{xcolor}
\usepackage{enumitem}

% --- Custom commands ---
\newcommand{\HH}[1]{H_{#1}}    % homology dimension
\newcommand{\PH}{\mathrm{PH}}
\newcommand{\VR}{\mathrm{VR}}
\newcommand{\Wass}{W_1}         % 1-Wasserstein distance
\newcommand{\pval}{\textit{p}}  % p-value formatting
\newcommand{\placeholder}[1]{\textbf{[#1]}}  % mark placeholders

% --- Title ---
\title{Persistent Homology of UK Socioeconomic Life-Course Trajectories}
\author{%
  [Author Names]\\
  \textit{[Affiliations]}
}
\date{March 2026}

% ============================================================
\begin{document}
\maketitle

\begin{abstract}
Longitudinal panel data captures the temporal structure of life-course dynamics, yet conventional sequence analysis methods treat trajectories as character strings, discarding their geometric structure in high-dimensional space.
We apply persistent homology (PH) to 27,280 employment-income trajectories drawn from Understanding Society (2009--2023), with cross-era validation on 8,509 trajectories from the British Household Panel Study (1991--2008).
Each trajectory is a sequence of nine states combining employment status with income band, embedded into 20-dimensional PCA space via $n$-gram frequency vectors and analysed using Vietoris--Rips persistent homology with maxmin landmarks.
We introduce a \textbf{Markov memory ladder}---a battery of five null models of escalating structural complexity---that characterises the generating process underlying the observed topology.

Seven distinct mobility regimes emerge via Gaussian mixture models optimised by BIC.
Total-persistence tests suggest that temporal ordering creates topological structure (order-shuffle $\HH{0}$ $\pval < 0.005$) while first-order Markov dynamics fully account for the topology (Markov-1 $\HH{0}$ $\pval = 1.000$).
However, upgrading the test statistic from total persistence to Wasserstein diagram-level distances reverses the Markov finding.
Under Wasserstein testing (100 permutations, 2,000 landmarks), the Markov-1 null is rejected for $\HH{0}$ ($\pval = 0.002$): observed persistence diagrams differ significantly in shape from Markov-generated diagrams, even though Markov surrogates produce more total persistence.
Markov-2 is \emph{not} rejected under Wasserstein ($\pval = 0.546$), localising the non-Markovian structure to second-order memory.
Order-shuffle $\HH{0}$ is borderline under Wasserstein ($\pval = 0.058$), label shuffle serves as a negative control ($\pval = 0.452$), and $\HH{1}$ (loops) remains non-significant under all null types.
This discrepancy between scalar and diagram-level tests is itself a methodological finding: total persistence masks structural differences that Wasserstein distances detect.

These findings replicate on BHPS-era trajectories (1991--2008): order-shuffle $\HH{0}$ $\pval = 0.000$, Markov-1 $\HH{0}$ $\pval = 0.000$---with the BHPS data additionally rejecting Markov-1 $\HH{1}$ ($\pval = 0.000$).
Stratified Wasserstein distances reveal significant topological differences by gender, parental NS-SEC, and birth cohort; 30/50 tests survive Benjamini--Hochberg FDR correction at $q < 0.05$.
Overlapping 10-year career-phase windows confirm extreme regime stickiness: among 7,453 individuals starting in disadvantaged regimes, only 416 (5.6\%) escape to advantaged regimes.

\medskip
\noindent\textbf{Keywords:} persistent homology, topological data analysis, social mobility, life-course trajectories, Markov models, regime switching, Wasserstein distance

\medskip
\noindent\textbf{JEL Codes:} C14, C38, J62, D63
\end{abstract}

\newpage
\tableofcontents
\newpage

% ============================================================
% Body content converted from markdown draft
% To regenerate: python scripts/md2latex.py --draft drafts/v6-2026-03.md --output latex/body.tex
\input{body}

% ============================================================
% BIBLIOGRAPHY
\nocite{*}  % Include all references while citations are converted from markdown
\bibliography{references}


% ============================================================
\appendix

\section{\texorpdfstring{$\HH{0}$}{H0} Merge Tree Analysis}
\label{app:merge}

\placeholder{Appendix A}

\section{Landmark Sensitivity}
\label{app:landmarks}

\placeholder{Appendix B}

\end{document}
